---
title: "Lab 5: Bikeshare SpaceTime Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Johnny"
date: today
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# INTRODUCTION

## The Rebalancing Challenge in Philadelphia

Philadelphia's Indego bike share system faces the same operational challenge as every bike share system: **rebalancing bikes to meet anticipated demand**.

Imagine you're an Indego operations manager at 6:00 AM on a Monday morning. You have: - 200 stations across Philadelphia - Limited trucks and staff for moving bikes - 2-3 hours before morning rush hour demand peaks - **The question:** Which stations will run out of bikes by 8:30 AM?

This lab will teach you to build predictive models that forecast bike share demand across **space** (different stations) and **time** (different hours) to help solve this operational problem.

```{r setup}
#| message: false
#| warning: false
#| include: false
library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
Sys.setlocale("LC_TIME", "C")
options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

```{r themes}
#| include: false
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}
#| include: false
census_api_key(Sys.getenv("CENSUS_API_KEY"))

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```

```{r load_indego}
#| include: false
# Read Q1 2025 data
indego <- read_csv(here("data/indego-trips-2025-q1.csv"))

# Quick look at the data
glimpse(indego)
```

```{r explore_data}
#| include: false
# How many trips?
cat("Total trips in Q1 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

```{r create-time-bins}
#| include: false
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

```{r trips_over_time}
#| message: false
#| warning: false
#| include: false
# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q1 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

```{r}
#| include: false
fly_eagles_fly <- daily_trips %>% filter(date == "2025-02-14")

typical_boring_friday <- indego %>%
  filter(dotw == "Fri", date != "2025-02-14") %>%
  group_by(date) %>%
  summarize(trips = n()) %>%
  summarize(avg_friday_trips = mean(trips))

print(fly_eagles_fly)
print(typical_boring_friday)

```

```{r hourly_patterns}
#| include: false
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

```{r top_stations}
#| include: false
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r load_census}
#| message: false
#| warning: false
#| include: false

# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```

```{r map_philly}
#| include: false
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

```{r join_census_to_stations}
#| include: false
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

```{r}
#| include: false
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

```{r get_weather}
#| include: false
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-01-01",
  date_end = "2025-03-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

```{r visualize-weather}
#| message: false
#| warning: false
#| include: false
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

```{r aggregate-trips}
#| message: false
#| warning: false
#| include: false
# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

```{r complete-panel}
#| include: false
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

```{r add_time_features}
#| include: false
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

```{r join_weather}
#| include: false
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

```{r create_lags}
#| include: false
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

```{r lag_correlations}
#| include: false
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

```{r temporal-split}
#| include: false
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

```{r model1}
#| include: false
# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

```{r model2}
#| include: false
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

```{r model3}
#| include: false
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

```{r model4}
#| include: false
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

```{r model5}
#| include: false
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

```{r calculate_mae}
#| include: false
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r compare_models}
#| include: false
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r obs_vs_pred}
#| message: false
#| warning: false
#| include: false
test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

```{r spatial_errors}
#| include: false
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))




# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

```{r temporal_errors}
#| include: false
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r errors_demographics}
#| message: false
#| warning: false
#| include: false
# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

## Part 1: Replicate with Different Quarter (alternately, you could do a longer time-span by merging multiple quarters together. I'm not picky about that)

1.  **Download data** for Q2, Q3, or Q4 2024 from: https://www.rideindego.com/about/data/

```{r load_indego_q2}
indego_q2 <- read_csv(here("data/indego-trips-2025-q2.csv"))
glimpse(indego_q2)
```

2.  **Adapt this code** to work with your quarter:

```{r explore_data_q2}
#| include: false
cat("Total trips in Q2 2025:", nrow(indego_q2), "\n")

cat("Date range:",
    min(mdy_hm(indego_q2$start_time)), "to",
    max(mdy_hm(indego_q2$start_time)), "\n")

cat("Unique start stations:", length(unique(indego_q2$start_station)), "\n")

table(indego_q2$trip_route_category)
table(indego_q2$passholder_type)
table(indego_q2$bike_type)
```

```{r create-time-bins_q2}
indego_q2 <- indego_q2 %>%
  mutate(
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    interval60 = floor_date(start_datetime, unit = "hour"),
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE, abbr = TRUE, locale = "C"),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

head(indego_q2 %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

#EDA OF INDEGO DATA \## Trips Over Time

```{r EDA_trips_over_time_q2}
# Daily trip counts
daily_q2_trips <- indego_q2 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_q2_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q2 2025",
    subtitle = "Spring demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

## Hourly Patterns

```{r EDA_hourly_patterns_q2}
# Average trips by hour and day type
hourly_q2_patterns <- indego_q2 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_q2_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

## Top Stations

```{r top_stations_q2}
# Most popular origin stations
top_q2_stations <- indego_q2 %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(
  top_q2_stations,
  caption = "Top 20 Indego Stations by Trip Origins",
  format.args = list(big.mark = ",")
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUT DATA IN CONTEXT

## Load Philadelphia Census Data

```{r load_census_q2}
#| include: false
philly_census_q2 <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)

glimpse(philly_census_q2)
```

## Map Philadelphia Context

```{r map_philly_q2}
ggplot() +
  geom_sf(data = philly_census_q2, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  geom_point(
    data = indego_q2,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

## Join Census Data to Stations

```{r join_census_to_stations_q2}
#| echo: false
# Create sf object for stations
stations_q2_sf <- indego_q2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_q2_census <- st_join(stations_q2_sf, philly_census_q2, left = TRUE) %>%
  st_drop_geometry()

# Prepare data for visualization
stations_q2_for_map <- indego_q2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_q2_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add census data back to trip data
indego_q2_census <- indego_q2 %>%
  left_join(
    stations_q2_census %>%
      select(start_station, Med_Inc, Percent_Taking_Transit,
             Percent_White, Total_Pop),
    by = "start_station"
  )

ggplot() +
  geom_sf(data = philly_census_q2, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  geom_point(
    data = stations_q2_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  geom_point(
    data = stations_q2_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme
```

## Dealing with missing data

```{r missing data_q2}
#| include: false
# Identify which stations to keep
valid_q2_stations <- stations_q2_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_q2_census <- indego_q2 %>%
  filter(start_station %in% valid_q2_stations) %>%
  left_join(
    stations_q2_census %>%
      select(start_station, Med_Inc, Percent_Taking_Transit,
             Percent_White, Total_Pop),
    by = "start_station"
  )
```

## Get Weather Data

```{r get_weather_q2}
#| include: false
weather_data_q2 <- riem_measures(
  station = "PHL",
  date_start = "2025-04-01",
  date_end = "2025-06-30"
)

weather_processed_q2 <- weather_data_q2 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,
    Precipitation = ifelse(is.na(p01i), 0, p01i),
    Wind_Speed = sknt
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

weather_complete_q2 <- weather_processed_q2 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

summary(weather_complete_q2 %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

```{r visualize-weather_q2}
ggplot(weather_complete_q2, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q2 2025",
    subtitle = "Spring to early summer transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips_q2}
#| message: false
#| warning: false
#| echo: false
# Count trips by station-hour
trips_q2_panel <- indego_q2_census %>%
  group_by(
    interval60, start_station, start_lat, start_lon,
    Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop
  ) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

nrow(trips_q2_panel)
length(unique(trips_q2_panel$start_station))
length(unique(trips_q2_panel$interval60))

hist(trips_q2_panel$Trip_Count)
```

## Create Complete Panel Structure

```{r complete-panel_q2}
#| echo: false
n_stations_q2 <- length(unique(trips_q2_panel$start_station))
n_hours_q2 <- length(unique(trips_q2_panel$interval60))
expected_rows_q2 <- n_stations_q2 * n_hours_q2

cat("Expected panel rows:", format(expected_rows_q2, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_q2_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows_q2 - nrow(trips_q2_panel), big.mark = ","), "\n")

study_q2_panel <- expand_grid(
  interval60 = unique(trips_q2_panel$interval60),
  start_station = unique(trips_q2_panel$start_station)
) %>%
  left_join(
    trips_q2_panel %>%
      select(interval60, start_station, Trip_Count),
    by = c("interval60", "start_station")
  ) %>%
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_q2_panel)
hist(study_q2_panel$Trip_Count)

station_q2_attributes <- trips_q2_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_q2_attributes)

study_q2_panel <- study_q2_panel %>%
  left_join(station_q2_attributes, by = "start_station")

glimpse(study_q2_panel)

cat("Complete panel rows:", format(nrow(study_q2_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features_q2}
#| include: false
study_q2_panel <- study_q2_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_q2_panel)
```

## Join Weather Data

```{r join_weather_q2}
#| include: false
study_q2_panel <- study_q2_panel %>%
  left_join(weather_complete_q2, by = "interval60")

glimpse(study_q2_panel)

summary(study_q2_panel %>% select(Trip_Count, Temperature, Precipitation))
```

# CREATE TEMPORAL LAG VARIABLES

```{r create_lags_q2}
#| include: false
study_q2_panel <- study_q2_panel %>%
  arrange(start_station, interval60)

study_q2_panel <- study_q2_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

study_q2_panel_complete <- study_q2_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:",
    format(nrow(study_q2_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations_q2}
#| echo: false
# Sample one station to visualize
example_q2_station <- study_q2_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_q2_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

```{r temporal-split_q2}
#| echo: false
early_q2_stations <- study_q2_panel_complete %>%
  filter(week < 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_q2_stations <- study_q2_panel_complete %>%
  filter(week >= 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

common_q2_stations <- intersect(early_q2_stations, late_q2_stations)

study_q2_panel_complete <- study_q2_panel_complete %>%
  filter(start_station %in% common_q2_stations)

train_q2 <- study_q2_panel_complete %>%
  filter(week < 23)

test_q2 <- study_q2_panel_complete %>%
  filter(week >= 23)

cat("Training observations:", format(nrow(train_q2), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test_q2), big.mark = ","), "\n")
cat("Training date range:", min(train_q2$date), "to", max(train_q2$date), "\n")
cat("Testing date range:", min(test_q2$date), "to", max(test_q2$date), "\n")
```

#BUILD PREDICTIVE MODELS

## Model 1: Baseline (Time + Weather)

```{r model1_q2}
#| include: false
# Create day-of-week factor with treatment coding
train_q2 <- train_q2 %>%
  mutate(
    dotw_simple_q2 = factor(
      dotw,
      levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )

# Set contrasts to treatment coding
contrasts(train_q2$dotw_simple_q2) <- contr.treatment(7)

model1_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation,
  data = train_q2
)

summary(model1_q2)
```

## Model 2: Add Temporal Lags

```{r model2_q2}
#| include: false
model2_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train_q2
)

summary(model2_q2)
```

## Model 3: Add Demographics

```{r model3_q2}
#| include: false
model3_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train_q2
)

summary(model3_q2)
```

## Model 4: Add Station Fixed Effects

```{r model4_q2}
#| include: false
model4_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train_q2
)

# Summary is long because of station dummies
cat("Model 4 R-squared:", summary(model4_q2)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4_q2)$adj.r.squared, "\n")
```

## Model 5: Add Rush Hour Interaction

```{r model5_q2}
#| include: false
model5_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,
  data = train_q2
)

cat("Model 5 R-squared:", summary(model5_q2)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5_q2)$adj.r.squared, "\n")
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae_q2}
#| echo: false
# Make sure test set has the same day-of-week factor used in training
test_q2 <- test_q2 %>%
  mutate(
    dotw_simple_q2 = factor(
      as.character(dotw),
      levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )

# Generate predictions on test set
test_q2 <- test_q2 %>%
  mutate(
    pred1_q2 = predict(model1_q2, newdata = test_q2),
    pred2_q2 = predict(model2_q2, newdata = test_q2),
    pred3_q2 = predict(model3_q2, newdata = test_q2),
    pred4_q2 = predict(model4_q2, newdata = test_q2),
    pred5_q2 = predict(model5_q2, newdata = test_q2)
  )

# Calculate MAE for each model
mae_q2_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test_q2$Trip_Count - test_q2$pred1_q2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred2_q2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred3_q2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred4_q2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred5_q2), na.rm = TRUE)
  )
)

kable(
  mae_q2_results,
  digits = 2,
  caption = "Mean Absolute Error by Model (Test Set, Q2 2025)",
  col.names = c("Model", "MAE (trips)")
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models_q2}
#| echo: false
ggplot(mae_q2_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison - Q2 2025",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

3.  **Compare results** to Q1 2025: \# Visualize Compare results \## MAE comparison

```{r compare_mae_q1_q2}
#| echo: false
# Combine MAE results from Q1 and Q2
mae_compare <- bind_rows(
  mae_results %>% mutate(Quarter = "Q1 2025"),
  mae_q2_results %>% mutate(Quarter = "Q2 2025")
)

ggplot(mae_compare, aes(x = Model, y = MAE, fill = Quarter)) +
  geom_col(position = "dodge", alpha = 0.9) +
  geom_text(
    aes(label = round(MAE, 2)),
    position = position_dodge(width = 0.9),
    vjust = -0.5,
    size = 3
  ) +
  scale_fill_manual(values = c(
    "Q1 2025" = "#B24A4A",
    "Q2 2025" = "#3182bd"
  )) +
  labs(
    title = "Model Performance Comparison: Q1 vs Q2 2025",
    subtitle = "Lower MAE indicates better prediction accuracy",
    x = "Model",
    y = "Mean Absolute Error (trips)",
    fill = "Quarter"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Daily trips comparison

```{r compare_daily_trip_q1_q2}
#| echo: false
daily_q1_compare <- daily_trips %>%
  arrange(date) %>%
  mutate(
    day_in_quarter = row_number(),
    Quarter = "Q1 2025"
  )

daily_q2_compare <- daily_q2_trips %>%
  arrange(date) %>%
  mutate(
    day_in_quarter = row_number(),
    Quarter = "Q2 2025"
  )

daily_compare <- bind_rows(daily_q1_compare, daily_q2_compare)

ggplot(daily_compare, aes(x = day_in_quarter, y = trips, color = Quarter)) +
  geom_line(linewidth = 1) +
  geom_smooth(se = FALSE, linetype = "dashed", linewidth = 0.9) +
  scale_color_manual(values = c(
    "Q1 2025" = "#B24A4A",
    "Q2 2025" = "#3182bd"
  )) +
  labs(
    title = "Daily Indego Ridership: Q1 vs Q2 2025",
    subtitle = "Comparing daily trip patterns across quarters",
    x = "Day in Quarter",
    y = "Daily Trips",
    color = "Quarter"
  ) +
  plotTheme
```

## Hourly pattern comparison

```{r compare_hourly_pattern_q1_q2}
#| echo: false
hourly_q1_compare <- hourly_patterns %>%
  mutate(Quarter = "Q1 2025")

hourly_q2_compare <- hourly_q2_patterns %>%
  mutate(Quarter = "Q2 2025")

hourly_compare <- bind_rows(hourly_q1_compare, hourly_q2_compare)

ggplot(hourly_compare, aes(x = hour, y = avg_trips, color = Quarter)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ day_type) +
  scale_color_manual(values = c(
    "Q1 2025" = "#B24A4A",
    "Q2 2025" = "#3182bd"
  )) +
  labs(
    title = "Average Hourly Ridership Patterns: Q1 vs Q2 2025",
    subtitle = "Comparing weekday and weekend demand profiles",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Quarter"
  ) +
  plotTheme
```

```         
-   How do MAE values compare? Why might they differ?
-   Are temporal patterns different (e.g., summer vs. winter)?
-   Which features are most important in your quarter?
```

MAE values are consistently higher in Q2 2025 than in Q1 2025. One likely reason is that overall bike-share demand is higher in Q2, so prediction errors measured in absolute trip counts also become larger. In addition, Q2 covers April to June, when warmer and more variable weather may lead to more complex travel behavior. Compared with the colder and more commute-driven patterns in Q1, Q2 likely includes more discretionary and weather-sensitive trips, which makes demand harder to predict and increases model error.

Q1 and Q2 show similar hourly structures: weekday demand peaks during the morning and evening commute periods, while weekend demand stays elevated from mid-day to early evening. However, the daily ridership comparison suggests a clearer difference across quarters. In Q1, daily trips generally increase over time, while Q2 shows a more fluctuating pattern, with alternating periods of increase and decline. A likely explanation is seasonal weather variation. Bike-share demand may be lower when temperatures are either too cold or too hot, and Q2 appears to be more sensitive to these changes. In particular, the drop in temperature during May may be associated with the first decline in trips, while the sustained temperature increase in June may help explain the second decline.

The most important features in my quarter are the temporal lag variables. This is clear from the large drop in MAE from Model 1 to Model 2, after adding lag1Hour, lag3Hours, and lag1day. This suggests that recent station demand is the strongest predictor of future demand in Q2 2025.

## Part 2: Error Analysis

Analyze your model's errors in detail:

1.  **Spatial patterns:**

    -   Create error maps
    -   Identify neighborhoods with high errors
    -   Hypothesize why (missing features? different demand patterns?)

2.  **Temporal patterns:**

    -   When are errors highest?
    -   Do certain hours/days have systematic under/over-prediction?
    -   Are there seasonal patterns?

3.  **Demographic patterns:**

    -   Relate errors to census characteristics
    -   Are certain communities systematically harder to predict?
    -   What are the equity implications?

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

```{r obs_vs_pred_q2}
#| echo: false
#| message: false
#| warning: false
# Use Model 2 for error analysis to stay consistent with the template
test_q2 <- test_q2 %>%
  mutate(
    error_q2 = Trip_Count - pred2_q2,
    abs_error_q2 = abs(error_q2),
    time_of_day_q2 = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

ggplot(test_q2, aes(x = Trip_Count, y = pred2_q2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day_q2) +
  labs(
    title = "Observed vs. Predicted Bike Trips - Q2 2025",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

## Spatial Error Patterns

```{r spatial_errors_q2}
#| echo: false
# Calculate station-level errors using Model 2
station_q2_errors <- test_q2 %>%
  filter(!is.na(pred2_q2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2_q2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census_q2, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_q2_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1
  ) +
  labs(title = "Prediction Errors - Q2 2025") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census_q2, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_q2_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1
  ) +
  labs(title = "Average Demand - Q2 2025") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

(p1 | p2)
```

## Temporal Error Patterns

```{r temporal_errors_q2}
#| message: false
#| warning: false
#| echo: false
temporal_q2_errors <- test_q2 %>%
  group_by(time_of_day_q2, weekend) %>%
  summarize(
    MAE = mean(abs_error_q2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_q2_errors, aes(x = time_of_day_q2, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(
    "Weekday" = "#08519c",
    "Weekend" = "#6baed6"
  )) +
  labs(
    title = "Prediction Errors by Time Period - Q2 2025",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

```{r errors_demographics_q2}
#| message: false
#| warning: false
#| echo: false
# Join demographic data to station-level errors
station_q2_errors_demo <- station_q2_errors %>%
  left_join(
    station_q2_attributes %>%
      select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Errors vs. median income
p1 <- ggplot(station_q2_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(
    title = "Errors vs. Median Income",
    x = "Median Income",
    y = "MAE"
  ) +
  plotTheme

# Errors vs. transit usage
p2 <- ggplot(station_q2_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Errors vs. Transit Usage",
    x = "% Taking Transit",
    y = "MAE"
  ) +
  plotTheme

# Errors vs. racial composition
p3 <- ggplot(station_q2_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Errors vs. Race",
    x = "% White",
    y = "MAE"
  ) +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

Spatial patterns errors are concentrated in the central part of Philadelphia, where demand is also highest. This suggests that the model struggles most at the busiest stations. A likely reason is that these central stations have more complex demand patterns, including commuting, which are not fully captured by the current model.

Temporal patterns errors are highest during PM Rush, and they are also relatively high during AM Rush and Mid-Day, while overnight errors are lowest. This means the model performs worst when demand is high and changes quickly. The observed-versus-predicted plots also suggest some under-prediction at peak demand levels. Q2 likely has stronger seasonal and weather-related variation, which may further increase daytime prediction error.

Demographic prediction errors are associated with neighborhood characteristics. Errors tend to increase with median income and % White, and decrease as transit usage increases. This suggests that some communities may be systematically harder to predict, but these should be interpreted as associations rather than causal effects. The equity concern is that if errors are concentrated in certain communities, forecast-based service decisions may also become uneven.

## Part 3: Feature Engineering & model improvement

```{r feature_engineering_q2}
#| include: false
# Create feature-engineered panel from the complete Q2 panel
study_q2_panel_fe <- study_q2_panel %>%
  arrange(start_station, interval60) %>%
  group_by(start_station) %>%
  mutate(
    # Same hour last week = 24 hours * 7 days = 168 hours
    same_hour_last_week_q2 = lag(Trip_Count, 168)
  ) %>%
  ungroup() %>%
  mutate(
    # Perfect biking weather: 60-75F and no rain
    perfect_weather_q2 = ifelse(
      Temperature >= 60 & Temperature <= 75 & Precipitation == 0,
      1, 0
    )
  )

# Use Philadelphia City Hall as a proxy for Center City
center_city_q2 <- tibble(
  place = "Center City",
  lon = -75.1636,
  lat = 39.9526
) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(2272)

# Compute station distance to Center City in feet
station_q2_distance <- station_q2_attributes %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326) %>%
  st_transform(2272) %>%
  mutate(
    dist_center_city_q2 = as.numeric(st_distance(geometry, center_city_q2))
  ) %>%
  st_drop_geometry() %>%
  select(start_station, dist_center_city_q2)

# Join distance back to panel
study_q2_panel_fe <- study_q2_panel_fe %>%
  left_join(station_q2_distance, by = "start_station")

# Remove rows where same-hour-last-week is missing
study_q2_panel_fe <- study_q2_panel_fe %>%
  filter(!is.na(same_hour_last_week_q2))
```

```{r split_fe_q2}
#| include: false
early_q2_stations_fe <- study_q2_panel_fe %>%
  filter(week < 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_q2_stations_fe <- study_q2_panel_fe %>%
  filter(week >= 23) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

common_q2_stations_fe <- intersect(early_q2_stations_fe, late_q2_stations_fe)

study_q2_panel_fe <- study_q2_panel_fe %>%
  filter(start_station %in% common_q2_stations_fe)

train_q2_fe <- study_q2_panel_fe %>%
  filter(week < 23) %>%
  mutate(
    dotw_simple_q2 = factor(
      as.character(dotw),
      levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )

test_q2_fe <- study_q2_panel_fe %>%
  filter(week >= 23) %>%
  mutate(
    dotw_simple_q2 = factor(
      as.character(dotw),
      levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )
```

```{r model6_q2}
#| include: false
model6_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    same_hour_last_week_q2 + dist_center_city_q2 + perfect_weather_q2,
  data = train_q2_fe
)

summary(model6_q2)
```

```{r compare_model6_mae_q2}
#| echo: false
test_q2_fe <- test_q2_fe %>%
  mutate(
    pred2_base_q2 = predict(model2_q2, newdata = test_q2_fe),
    pred6_q2 = predict(model6_q2, newdata = test_q2_fe)
  )

mae_model2_base_q2 <- mean(abs(test_q2_fe$Trip_Count - test_q2_fe$pred2_base_q2), na.rm = TRUE)
mae_model6_q2 <- mean(abs(test_q2_fe$Trip_Count - test_q2_fe$pred6_q2), na.rm = TRUE)

mae_part3_q2 <- data.frame(
  Model = c("Model 2 Baseline", "Model 6 + New Features"),
  MAE = c(mae_model2_base_q2, mae_model6_q2)
)

kable(
  mae_part3_q2,
  digits = 3,
  caption = "MAE Comparison: Baseline vs. Feature-Engineered Model (Q2 2025)"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r compare_model6_plot_q2}
#| echo: false
ggplot(mae_part3_q2, aes(x = Model, y = MAE, fill = Model)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = round(MAE, 3)), vjust = -0.5) +
  scale_fill_manual(values = c(
    "Model 2 Baseline" = "#B24A4A",
    "Model 6 + New Features" = "#3182bd"
  )) +
  labs(
    title = "Feature Engineering Improvement - Q2 2025",
    subtitle = "Lower MAE indicates better prediction accuracy",
    x = NULL,
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(legend.position = "none")
```

```{r poisson_model_q2}
#| include: false
model6_pois_q2 <- glm(
  Trip_Count ~ as.factor(hour) + dotw_simple_q2 + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    same_hour_last_week_q2 + dist_center_city_q2 + perfect_weather_q2,
  data = train_q2_fe,
  family = poisson(link = "log")
)

summary(model6_pois_q2)
```

```{r poisson_mae_q2}
#| echo: false
test_q2_fe <- test_q2_fe %>%
  mutate(
    pred6_pois_q2 = predict(model6_pois_q2, newdata = test_q2_fe, type = "response")
  )

mae_model6_pois_q2 <- mean(abs(test_q2_fe$Trip_Count - test_q2_fe$pred6_pois_q2), na.rm = TRUE)

mae_poisson_compare_q2 <- data.frame(
  Model = c("Model 2 Baseline", "Model 6 Linear", "Model 6 Poisson"),
  MAE = c(mae_model2_base_q2, mae_model6_q2, mae_model6_pois_q2)
)

kable(
  mae_poisson_compare_q2,
  digits = 3,
  caption = "MAE Comparison: Baseline, Linear FE Model, and Poisson FE Model (Q2 2025)"
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

In this part, I added three new features to Model 2: same_hour_last_week_q2, dist_center_city_q2, and perfect_weather_q2. After adding these variables, the test-set MAE decreased from 0.690 to 0.673, which indicates a small improvement in predictive accuracy.

I chose same_hour_last_week_q2 because the model showed larger errors during peak periods, and demand at the same hour in the previous week may help capture recurring temporal patterns. I added dist_center_city_q2 because the spatial error maps suggested that both higher errors and higher demand were concentrated in the central city, so distance to Center City may help explain this spatial variation. I also added perfect_weather_q2 because ridership in Q2 appeared to be more sensitive to seasonal and weather conditions, with demand patterns moving together with changes in temperature.

These features improved model performance, but only slightly. Among them, same_hour_last_week_q2 and dist_center_city_q2 were more useful, while perfect_weather_q2 had a weaker effect. This suggests that the model is still missing other important factors, such as holidays and other contextual influences that affect ridership. In addition, the Poisson model did not improve performance and actually performed worse than Model 6.

## Part 4: Critical Reflection

Write 1-2 paragraphs addressing:

1.  **Operational implications:**
    -   Is your final MAE "good enough" for Indego to use?
    -   When do prediction errors cause problems for rebalancing?
    -   Would you recommend deploying this system? Under what conditions?
2.  **Equity considerations:**
    -   Do prediction errors disproportionately affect certain neighborhoods?
    -   Could this system worsen existing disparities in bike access?
    -   What safeguards would you recommend?
3.  **Model limitations:**
    -   What patterns is your model missing?
    -   What assumptions might not hold in real deployment?
    -   How would you improve this with more time/data?

My final model is useful, but it is not accurate enough to support fully automated rebalancing on its own. An MAE of about 0.673 suggests that it can help Indego identify overall demand patterns, but prediction errors would still create operational problems when they occur at high-demand stations or during peak hours. For that reason, this model is better understood as an initial decision-support tool rather than a replacement for human rebalancing decisions.

At the same time, the model also has equity and methodological limitations. My results suggest that prediction errors are not evenly distributed across neighborhoods, so direct deployment could unintentionally worsen existing inequalities in bike access. To reduce that risk, Indego should monitor errors by neighborhood, provide targeted safeguards for underserved areas, and avoid relying entirely on automated predictions. The current model is still a preliminary version.With more time and data, I would improve the model in three ways. First, I would add richer temporal and event variables, such as holidays, school calendars, and major events, because the current model still shows large variation across certain dates. Second, I would include more detailed spatial and station-level information, such as station type, nearby points of interest, and bike network conditions, to better explain why high-demand central areas still have relatively large errors. Finally, I would add more demographic and socioeconomic variables, such as income, race, and education, because the error analysis suggests that these neighborhood characteristics are associated with variation in prediction errors.

------------------------------------------------------------------------
