---
title: "CPLN 5920 Final Project: EDA"
format: html
editor: visual
---

## Start with the outcome variable (filings)
1. How skewed is it?
2. How many zeros?
3. Overdispersion?

```{r setup, include=FALSE}
#| message: false
#| warning: false

# Load required packages
library(tidyverse)      # Data manipulation
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)       # Classification intervals
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidycensus)


# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

```{r load data set}
evict <- read_csv("data/eviction.csv")

glimpse(evict)

evict <- evict %>%
  mutate(
    GEOID = as.character(GEOID),
    month_date = my(month),   # "03/2023" → 2023-03-01
    year = year(month_date),
    month_num = month(month_date)
  ) %>%
  dplyr::select(-last_updated)

```

<hr>
## Distribution

```{r}
ggplot(evict, aes(x = filings_2020)) +
  geom_histogram(bins = 50, fill = "#1f4e79", color = "white", linewidth = 0.2)

```
Eviction filings are highly skewed and there are a lot of zeros.

<hr>
## Time Trends
1. Upward/downward trends
2. Seasonal patterns

```{r}
evict %>%
  group_by(month_date) %>%
  summarise(avg_filings = mean(filings_2020, na.rm = TRUE)) %>%
  ggplot(aes(x = month_date, y = avg_filings)) +
  geom_line()

```
There is no strong long-term trend in the filing data and very high month-to-month volatility.
<hr>
## Spatial Inequality
1. Top & bottom tracts
2. top 10 tracts by absolute filing numbers and concentration
3. spatial pattern maps

```{r}
evict %>%
  group_by(GEOID) %>%
  summarise(avg_filings = mean(filings_2020, na.rm = TRUE)) %>%
  ggplot(aes(x = avg_filings)) +
  geom_histogram(bins = 50, fill = "#1f4e79", color = "white", linewidth = 0.2)

```

```{r}
top10 <- evict %>%
  group_by(GEOID) %>%
  summarise(total = sum(filings_2020, na.rm = TRUE)) %>%
  arrange(desc(total)) %>%
  head(10) # Top 10 census tracts with the most filings

top10

# Pull renter households and housing cost variables
acs_housing <- get_acs(
  geography = "tract",
  variables = c(
    renters = "B25003_003",
    median_gross_rent = "B25064_001",
    median_household_income = "B19013_001",
    renter_hh_total = "B25070_001",
    rent_30_34 = "B25070_007",
    rent_35_39 = "B25070_008",
    rent_40_49 = "B25070_009",
    rent_50_plus = "B25070_010",
    rent_not_computed = "B25070_011"
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2023,
  geometry = FALSE
)

renters <- acs_housing %>%
  dplyr::select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    renter_hh_computed = renter_hh_total - rent_not_computed,
    rent_burden_pct = if_else(
      renter_hh_computed > 0,
      ((rent_30_34 + rent_35_39 + rent_40_49 + rent_50_plus) / renter_hh_computed) * 100,
      NA_real_
    )
  ) %>%
  dplyr::select(
    GEOID,
    renters,
    median_gross_rent,
    median_household_income,
    rent_burden_pct
  )

# Join to dataset
evict <- evict %>%
  left_join(renters, by = "GEOID")

# Normaliza filings
evict <- evict %>%
  mutate(
    filings_per_1k_renters = (filings_2020 / renters) * 1000
  )

# Compute tract-level average rate
evict_clean_rate <- evict %>%
  filter(!is.na(renters) & renters > 0)

evict_clean_rate <- evict_clean_rate %>%
  filter(renters > 100)   # adjust threshold

top10_rate <- evict_clean_rate %>%
  group_by(GEOID) %>%
  summarise(
    avg_rate = mean(filings_per_1k_renters, na.rm = TRUE),
    total_filings = sum(filings_2020, na.rm = TRUE),
    renters = first(renters),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_rate)) %>%
  slice(1:10)

top10_rate

# Find common tracts as policy hotspots
hotspot <- intersect(top10$GEOID, top10_rate$GEOID)
hotspot

```

```{r spatial-pattern-maps, message=FALSE, warning=FALSE}
# Build tract-level averages
tract_summary <- evict %>%
  filter(!is.na(GEOID)) %>%
  group_by(GEOID) %>%
  summarise(
    avg_filings = mean(filings_2020, na.rm = TRUE),
    avg_rate_1k = mean(filings_per_1k_renters, na.rm = TRUE),
    total_filings = sum(filings_2020, na.rm = TRUE),
    renters = first(renters),
    .groups = "drop"
  ) %>%
  filter(!is.na(avg_filings), !is.na(avg_rate_1k))

# Download Philadelphia tract geometry
phl_tracts <- get_acs(
  geography = "tract",
  variables = "B01003_001",
  state = "PA",
  county = "Philadelphia",
  year = 2023,
  geometry = TRUE
) %>%
  dplyr::select(GEOID, geometry)

# Join summary data to tract geometry
tract_map <- phl_tracts %>%
  left_join(tract_summary, by = "GEOID")

# Shared orange color palette
orange_low  <- "#fff5eb"
orange_mid  <- "#fdbb84"
orange_high <- "#d94801"

# Map 1: Average filings per tract
map_avg_filings <- ggplot(tract_map) +
  geom_sf(aes(fill = avg_filings), color = NA) +
  scale_fill_gradientn(
    colours = c(orange_low, orange_mid, orange_high),
    name = "Avg.\nfilings",
    na.value = "grey90"
  ) +
  labs(title = "Average Eviction Filings per Tract") +
  theme_void() +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

# Map 2: Filings per 1,000 renters
map_rate <- ggplot(tract_map) +
  geom_sf(aes(fill = avg_rate_1k), color = NA) +
  scale_fill_gradientn(
    colours = c(orange_low, orange_mid, orange_high),
    name = "Filings per\n1,000 renters",
    na.value = "grey90"
  ) +
  labs(title = "Eviction Filings per 1,000 Renters") +
  theme_void() +
  theme(
    plot.title = element_text(size = 11, face = "bold"),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )

# Combine maps side by side
map_avg_filings + map_rate +
  plot_annotation(
    title = "Spatial Patterns of Eviction Burden and Renter-Adjusted Risk",
    theme = theme(
      plot.title = element_text(size = 13, face = "bold", hjust = 0.5)
    )
  )
```

There are a few tracts that have very high filing records. By finding the top 10 census tracts with highest filing records and top 10 census tracts with highest filing record concentration, we identify two high risk census tracts that appear in both lists as rental assistance policy hotspots.
<hr>
## Dynamics
1. Create multiple lag variables (1 month lag, 1 quarter lag)
2. Visualize Changes

```{r}
evict <- evict %>%
  arrange(GEOID, month_date) %>%
  group_by(GEOID)

# 1 month lag
evict <- evict %>%
  mutate(
    lag_1m = lag(filings_2020, 1)
  )

# Quarter lag
evict <- evict %>%
  mutate(
    lag_3m = lag(filings_2020, 3)
  )

# Create change variables
evict <- evict %>%
  mutate(
    change_1m = filings_2020 - lag_1m,
    change_3m = filings_2020 - lag_3m
  )

# Visualize 1 month change
ggplot(evict, aes(x = change_1m)) +
  geom_histogram(bins = 50, fill = "#1f4e79", color = "white", linewidth = 0.2) +
  labs(title = "Monthly Change in Filings")

# Visualize quarterly change
ggplot(evict, aes(x = change_3m)) +
  geom_histogram(bins = 50, fill = "#1f4e79", color = "white", linewidth = 0.2) +
  labs(title = "Quarterly Change in Filings")

```
Both monthly and quarterly change have symmetrical distribution. Eviction filings are not consistently increasing or decreasing. Changes in eviction activity fluctuate around a stable baseline rather than showing persistent growth or decline. This suggests that recent lag components of filings are strong predictors for future risks.

<hr>
## Race/Demographic Patterns

```{r}
evict %>%
  group_by(racial_majority) %>%
  summarise(avg_filings = mean(filings_2020, na.rm = TRUE))

```
Black-majority neighborhoods have substantially higher average eviction filings. Even after adjusting for renter population, eviction activity remains higher in Black-majority neighborhoods, suggesting structural disparities in housing stability.
<hr>

## Modeling Comparison
1. Keep both monthly and quarterly forecasting results
2. Compare baseline and progressive negative binomial models at both time scales
3. Summarize which models help prediction and which help interpretation

```{r modeling-helpers}
library(broom)

score_model <- function(actual, predicted) {
  tibble(
    MAE = mean(abs(actual - predicted), na.rm = TRUE),
    RMSE = sqrt(mean((actual - predicted)^2, na.rm = TRUE)),
    Correlation = cor(actual, predicted, use = "complete.obs")
  )
}

classification_scores <- function(data, truth, pred) {
  tp <- sum(data[[truth]] == 1 & data[[pred]] == 1, na.rm = TRUE)
  fp <- sum(data[[truth]] == 0 & data[[pred]] == 1, na.rm = TRUE)
  fn <- sum(data[[truth]] == 1 & data[[pred]] == 0, na.rm = TRUE)
  tn <- sum(data[[truth]] == 0 & data[[pred]] == 0, na.rm = TRUE)

  tibble(
    Accuracy = (tp + tn) / (tp + tn + fp + fn),
    Precision = ifelse(tp + fp == 0, NA, tp / (tp + fp)),
    Recall = ifelse(tp + fn == 0, NA, tp / (tp + fn)),
    F1 = ifelse(
      (tp + fp) == 0 | (tp + fn) == 0,
      NA_real_,
      2 * ((tp / (tp + fp)) * (tp / (tp + fn))) / ((tp / (tp + fp)) + (tp / (tp + fn)))
    )
  )
}

evaluate_models <- function(data, split_name, model_specs, baseline_var) {
  bind_rows(lapply(names(model_specs), function(model_name) {
    spec <- model_specs[[model_name]]

    predicted <- if (!is.null(spec$type) && spec$type == "baseline") {
      pmax(data[[baseline_var]], 0)
    } else {
      predict(spec$model, newdata = data, type = "response")
    }

    predicted <- pmax(predicted, 0)

    score_model(data$filings_next, predicted) %>%
      mutate(model = model_name, split = split_name)
  }))
}

```

### Monthly Models

```{r create-monthly-model-data}
monthly_model_data <- evict %>%
  ungroup() %>%
  arrange(GEOID, month_date) %>%
  group_by(GEOID) %>%
  mutate(
    filings_next = lead(filings_2020, 1),
    filings_next_per_1k_renters = lead(filings_per_1k_renters, 1),
    lag_2m = lag(filings_2020, 2),
    lag_12m = lag(filings_2020, 12),
    rolling_mean_3m = (lag_1m + lag_2m + lag_3m) / 3
  ) %>%
  ungroup() %>%
  mutate(
    racial_majority = factor(racial_majority),
    month_factor = factor(month_num, levels = 1:12, labels = month.abb)
  ) %>%
  filter(
    !is.na(filings_next),
    !is.na(lag_1m),
    !is.na(lag_2m),
    !is.na(lag_3m),
    !is.na(lag_12m),
    !is.na(renters),
    !is.na(median_gross_rent),
    !is.na(median_household_income),
    !is.na(rent_burden_pct),
    renters > 0
  )

monthly_split_lookup <- tibble(
  month_date = sort(unique(monthly_model_data$month_date))
) %>%
  mutate(
    data_split = case_when(
      month_date <= ymd("2025-02-01") ~ "Train",
      month_date <= ymd("2025-08-01") ~ "Validation",
      TRUE ~ "Test"
    )
  )

monthly_model_data <- monthly_model_data %>%
  left_join(monthly_split_lookup, by = "month_date")

monthly_train <- monthly_model_data %>% filter(data_split == "Train")
monthly_validation <- monthly_model_data %>% filter(data_split == "Validation")
monthly_test <- monthly_model_data %>% filter(data_split == "Test")
```

```{r fit-monthly-models}
monthly_model_1 <- MASS::glm.nb(
  filings_next ~ lag_1m + offset(log(renters)),
  data = monthly_train
)

monthly_model_2 <- MASS::glm.nb(
  filings_next ~ lag_1m + lag_3m + change_1m + rolling_mean_3m + offset(log(renters)),
  data = monthly_train
)

monthly_model_3 <- MASS::glm.nb(
  filings_next ~ lag_1m + lag_3m + change_1m + rolling_mean_3m +
    filings_avg + filings_avg_prepandemic_baseline +
    offset(log(renters)),
  data = monthly_train
)

monthly_model_4 <- MASS::glm.nb(
  filings_next ~ lag_1m + lag_3m + change_1m + rolling_mean_3m +
    filings_avg + filings_avg_prepandemic_baseline +
    racial_majority + month_factor +
    offset(log(renters)),
  data = monthly_train
)

monthly_model_5 <- MASS::glm.nb(
  filings_next ~ lag_1m + lag_3m + change_1m + rolling_mean_3m +
    filings_avg + filings_avg_prepandemic_baseline +
    racial_majority + month_factor +
    median_gross_rent + median_household_income + rent_burden_pct +
    offset(log(renters)),
  data = monthly_train
)

monthly_enhanced_baseline <- lm(
  filings_next ~ lag_1m + lag_12m,
  data = monthly_train
)

monthly_rolling_baseline <- lm(
  filings_next ~ lag_1m + lag_12m + rolling_mean_3m,
  data = monthly_train
)

monthly_model_specs <- list(
  "Model 0: Baseline" = list(type = "baseline"),
  "Model 0b: + lag_12m baseline" = list(model = monthly_enhanced_baseline),
  "Model 0c: + rolling_mean_3m baseline" = list(model = monthly_rolling_baseline),
  "Model 1: lag_1m" = list(model = monthly_model_1),
  "Model 2: + dynamic terms" = list(model = monthly_model_2),
  "Model 3: + historical baselines" = list(model = monthly_model_3),
  "Model 4: + race and seasonality" = list(model = monthly_model_4),
  "Model 5: + rent and income" = list(model = monthly_model_5)
)
```

```{r compare-monthly-models}
monthly_scores <- bind_rows(
  evaluate_models(monthly_validation, "Validation", monthly_model_specs, "lag_1m"),
  evaluate_models(monthly_test, "Test", monthly_model_specs, "lag_1m")
) %>%
  mutate(scale = "Monthly")

monthly_scores %>%
  mutate(across(c(MAE, RMSE, Correlation), round, 3)) %>%
  dplyr::select(scale, split, model, MAE, RMSE, Correlation) %>%
  kable(caption = "Monthly Forecast Performance as Variables Are Added Step by Step") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r monthly-coefficients}
tidy(monthly_model_5, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value) %>%
  arrange(desc(estimate)) %>%
  mutate(across(c(estimate, conf.low, conf.high, p.value), round, 3)) %>%
  head(15) %>%
  kable(
    caption = "Monthly Negative Binomial Model Coefficients (Incidence Rate Ratios)",
    col.names = c("Term", "IRR", "Lower CI", "Upper CI", "p-value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r monthly-high-risk}
monthly_risk_threshold <- quantile(
  monthly_train$filings_next_per_1k_renters,
  probs = 0.90,
  na.rm = TRUE
)

monthly_test_predictions <- monthly_test %>%
  mutate(
    pred_next = predict(monthly_model_5, newdata = monthly_test, type = "response"),
    pred_next_per_1k_renters = (pred_next / renters) * 1000,
    high_risk_true = if_else(filings_next_per_1k_renters >= monthly_risk_threshold, 1, 0),
    high_risk_pred = if_else(pred_next_per_1k_renters >= monthly_risk_threshold, 1, 0)
  )

monthly_risk_scores <- classification_scores(monthly_test_predictions, "high_risk_true", "high_risk_pred")

monthly_risk_scores %>%
  mutate(across(everything(), round, 3)) %>%
  kable(caption = "Monthly High-Risk Classification Performance on Test Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Quarterly Models

```{r create-quarterly-model-data}
quarterly_model_data <- evict %>%
  ungroup() %>%
  mutate(
    quarter_date = floor_date(month_date, unit = "quarter"),
    quarter_label = paste0(year(quarter_date), " Q", quarter(quarter_date))
  ) %>%
  group_by(GEOID, quarter_date, quarter_label, racial_majority) %>%
  summarise(
    current_filings = sum(filings_2020, na.rm = TRUE),
    filings_avg = first(filings_avg),
    filings_avg_prepandemic_baseline = first(filings_avg_prepandemic_baseline),
    renters = first(renters),
    median_gross_rent = first(median_gross_rent),
    median_household_income = first(median_household_income),
    rent_burden_pct = first(rent_burden_pct),
    months_in_quarter = n(),
    .groups = "drop"
  ) %>%
  filter(months_in_quarter == 3) %>%
  arrange(GEOID, quarter_date) %>%
  group_by(GEOID) %>%
  mutate(
    filings_next = lead(current_filings, 1),
    filings_next_per_1k_renters = lead((current_filings / renters) * 1000, 1),
    lag_1q = lag(current_filings, 1),
    lag_2q = lag(current_filings, 2),
    change_1q = current_filings - lag_1q,
    rolling_mean_2q = (lag_1q + lag_2q) / 2
  ) %>%
  ungroup() %>%
  mutate(
    racial_majority = factor(racial_majority),
    quarter_factor = factor(quarter(quarter_date), levels = 1:4, labels = c("Q1", "Q2", "Q3", "Q4"))
  ) %>%
  filter(
    !is.na(filings_next),
    !is.na(lag_1q),
    !is.na(lag_2q),
    !is.na(renters),
    !is.na(median_gross_rent),
    !is.na(median_household_income),
    !is.na(rent_burden_pct),
    renters > 0
  )

quarterly_split_lookup <- tibble(
  quarter_date = sort(unique(quarterly_model_data$quarter_date))
) %>%
  mutate(
    data_split = case_when(
      quarter_date <= ymd("2024-10-01") ~ "Train",
      quarter_date <= ymd("2025-07-01") ~ "Validation",
      TRUE ~ "Test"
    )
  )

quarterly_model_data <- quarterly_model_data %>%
  left_join(quarterly_split_lookup, by = "quarter_date")

quarterly_train <- quarterly_model_data %>% filter(data_split == "Train")
quarterly_validation <- quarterly_model_data %>% filter(data_split == "Validation")
quarterly_test <- quarterly_model_data %>% filter(data_split == "Test")
```

```{r fit-quarterly-models}
quarterly_model_1 <- MASS::glm.nb(
  filings_next ~ lag_1q + offset(log(renters)),
  data = quarterly_train
)

quarterly_model_2 <- MASS::glm.nb(
  filings_next ~ lag_1q + lag_2q + change_1q + rolling_mean_2q + offset(log(renters)),
  data = quarterly_train
)

quarterly_model_3 <- MASS::glm.nb(
  filings_next ~ lag_1q + lag_2q + change_1q + rolling_mean_2q +
    filings_avg + filings_avg_prepandemic_baseline +
    offset(log(renters)),
  data = quarterly_train
)

quarterly_model_4 <- MASS::glm.nb(
  filings_next ~ lag_1q + lag_2q + change_1q + rolling_mean_2q +
    filings_avg + filings_avg_prepandemic_baseline +
    racial_majority + quarter_factor +
    offset(log(renters)),
  data = quarterly_train
)

quarterly_model_5 <- MASS::glm.nb(
  filings_next ~ lag_1q + lag_2q + change_1q + rolling_mean_2q +
    filings_avg + filings_avg_prepandemic_baseline +
    racial_majority + quarter_factor +
    median_gross_rent + median_household_income + rent_burden_pct +
    offset(log(renters)),
  data = quarterly_train
)

quarterly_model_specs <- list(
  "Model 0: Baseline" = list(type = "baseline"),
  "Model 1: lag_1q" = list(model = quarterly_model_1),
  "Model 2: + dynamic terms" = list(model = quarterly_model_2),
  "Model 3: + historical baselines" = list(model = quarterly_model_3),
  "Model 4: + race and seasonality" = list(model = quarterly_model_4),
  "Model 5: + rent and income" = list(model = quarterly_model_5)
)
```

```{r compare-quarterly-models}
quarterly_scores <- bind_rows(
  evaluate_models(quarterly_validation, "Validation", quarterly_model_specs, "lag_1q"),
  evaluate_models(quarterly_test, "Test", quarterly_model_specs, "lag_1q")
) %>%
  mutate(scale = "Quarterly")

quarterly_scores %>%
  mutate(across(c(MAE, RMSE, Correlation), round, 3)) %>%
  dplyr::select(scale, split, model, MAE, RMSE, Correlation) %>%
  kable(caption = "Quarterly Forecast Performance as Variables Are Added Step by Step") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r quarterly-coefficients}
tidy(quarterly_model_5, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value) %>%
  arrange(desc(estimate)) %>%
  mutate(across(c(estimate, conf.low, conf.high, p.value), round, 3)) %>%
  head(15) %>%
  kable(
    caption = "Quarterly Negative Binomial Model Coefficients (Incidence Rate Ratios)",
    col.names = c("Term", "IRR", "Lower CI", "Upper CI", "p-value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r quarterly-high-risk}
quarterly_risk_threshold <- quantile(
  quarterly_train$filings_next_per_1k_renters,
  probs = 0.90,
  na.rm = TRUE
)

quarterly_test_predictions <- quarterly_test %>%
  mutate(
    pred_next = predict(quarterly_model_5, newdata = quarterly_test, type = "response"),
    pred_next_per_1k_renters = (pred_next / renters) * 1000,
    high_risk_true = if_else(filings_next_per_1k_renters >= quarterly_risk_threshold, 1, 0),
    high_risk_pred = if_else(pred_next_per_1k_renters >= quarterly_risk_threshold, 1, 0)
  )

quarterly_risk_scores <- classification_scores(quarterly_test_predictions, "high_risk_true", "high_risk_pred")

quarterly_risk_scores %>%
  mutate(across(everything(), round, 3)) %>%
  kable(caption = "Quarterly High-Risk Classification Performance on Test Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r quarterly-top-risk-tracts}
quarterly_latest_forecast <- quarterly_test_predictions %>%
  filter(quarter_date == max(quarter_date, na.rm = TRUE)) %>%
  arrange(desc(pred_next_per_1k_renters)) %>%
  transmute(
    GEOID,
    racial_majority,
    quarter_label,
    predicted_filings_next_quarter = round(pred_next, 2),
    predicted_rate_next_quarter = round(pred_next_per_1k_renters, 2),
    predicted_high_risk = high_risk_pred
  )

quarterly_latest_forecast %>%
  slice(1:15) %>%
  kable(caption = "Top 15 Tracts by Predicted Next-Quarter Risk") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Monthly vs. Quarterly Summary

```{r overall-comparison-table}
overall_comparison <- bind_rows(
  monthly_scores %>%
    filter(model %in% c("Model 0: Baseline", "Model 0b: + lag_12m baseline", "Model 0c: + rolling_mean_3m baseline", "Model 5: + rent and income")) %>%
    mutate(scale = "Monthly"),
  quarterly_scores %>%
    filter(model %in% c("Model 0: Baseline", "Model 5: + rent and income")) %>%
    mutate(scale = "Quarterly")
) %>%
  mutate(across(c(MAE, RMSE, Correlation), round, 3))

overall_comparison %>%
  dplyr::select(scale, split, model, MAE, RMSE, Correlation) %>%
  kable(caption = "Overall Comparison of Monthly and Quarterly Forecasting Results") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r overall-comparison-plot}
overall_comparison_plot <- overall_comparison %>%
  mutate(
    model_short = case_when(
      model == "Model 0: Baseline" ~ "M0\nLag only",
      model == "Model 0b: + lag_12m baseline" ~ "M0b\n+ lag 12m",
      model == "Model 0c: + rolling_mean_3m baseline" ~ "M0c\n+ rolling mean",
      model == "Model 5: + rent and income" ~ "M5\nFull NB",
      TRUE ~ model
    )
  )

ggplot(overall_comparison_plot, aes(x = model_short, y = RMSE, fill = split)) +
  geom_col(position = "dodge", color = "white", linewidth = 0.2) +
  facet_wrap(~ scale, scales = "free_x") +
  scale_fill_manual(values = c("Validation" = "#6f8fb3", "Test" = "#17365d")) +
  labs(
    title = "Monthly vs. Quarterly Out-of-Sample Forecast Accuracy",
    subtitle = "Lower RMSE indicates better predictive performance",
    x = NULL,
    y = "RMSE"
  ) +
  theme(
    axis.text.x = element_text(size = 9),
    strip.text = element_text(face = "bold")
  )
```

The model comparison shows that monthly forecasting performs better than quarterly forecasting for predicting future eviction filings. The strongest predictive model is the enhanced monthly baseline, which combines last month’s filings, filings from the same month in the previous year, and a three-month rolling average. This suggests that eviction filings are driven primarily by short-term persistence and recurring seasonal patterns. In contrast, the richer negative binomial models with rent, income, demographic, and seasonal controls do not improve out-of-sample prediction accuracy, although they remain useful for interpreting structural differences in eviction risk across tracts.




































