---
title: "CPLN 5920 Final Project: Eviction Filing Prediction"
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
editor: visual
execute:
  warning: false
  message: false
---

# Introduction

This report builds a monthly tract-level prediction workflow for Philadelphia eviction filings. The modeling goal is twofold:

1.  Predict next month's filing count for each census tract.
2.  Flag which tracts are likely to enter a high-risk state.

Following the logic from the EDA, I use monthly data as the main unit of analysis because it preserves short-term dynamics and supports early-warning forecasting.

```{r setup}
#| include: false
library(tidyverse)
library(lubridate)
library(sf)
library(tidycensus)
library(MASS)
library(knitr)
library(kableExtra)
library(patchwork)
library(viridis)
library(here)
library(broom)

options(scipen = 999)
set.seed(5920)

plotTheme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption = element_text(size = 8),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption = element_text(size = 8),
    legend.position = "right"
  )

score_count_model <- function(data, actual, predicted) {
  tibble(
    MAE = mean(abs(data[[actual]] - data[[predicted]]), na.rm = TRUE),
    RMSE = sqrt(mean((data[[actual]] - data[[predicted]])^2, na.rm = TRUE)),
    Correlation = cor(data[[actual]], data[[predicted]], use = "complete.obs")
  )
}

score_classification <- function(data, truth, pred) {
  tp <- sum(data[[truth]] == 1 & data[[pred]] == 1, na.rm = TRUE)
  fp <- sum(data[[truth]] == 0 & data[[pred]] == 1, na.rm = TRUE)
  fn <- sum(data[[truth]] == 1 & data[[pred]] == 0, na.rm = TRUE)
  tn <- sum(data[[truth]] == 0 & data[[pred]] == 0, na.rm = TRUE)

  precision <- ifelse(tp + fp == 0, NA, tp / (tp + fp))
  recall <- ifelse(tp + fn == 0, NA, tp / (tp + fn))
  accuracy <- (tp + tn) / (tp + tn + fp + fn)
  f1 <- ifelse(is.na(precision) | is.na(recall) | precision + recall == 0,
               NA,
               2 * precision * recall / (precision + recall))

  tibble(
    Accuracy = accuracy,
    Precision = precision,
    Recall = recall,
    F1 = f1
  )
}
```

```{r set-census-api-key}
#| include: false
if (Sys.getenv("CENSUS_API_KEY") != "") {
  census_api_key(Sys.getenv("CENSUS_API_KEY"), install = FALSE, overwrite = FALSE)
}
```

# Load and Prepare Data

```{r load-data}
evict_raw <- read_csv(here("data", "eviction.csv"), show_col_types = FALSE)

glimpse(evict_raw)

evict <- evict_raw %>%
  mutate(
    GEOID = as.character(GEOID),
    month_date = my(month),
    year = year(month_date),
    month_num = month(month_date),
    filings_2020 = as.numeric(filings_2020),
    filings_avg = as.numeric(filings_avg),
    filings_avg_prepandemic_baseline = as.numeric(filings_avg_prepandemic_baseline)
  ) %>%
  dplyr::select(-last_updated)

cat("Rows:", nrow(evict), "\n")
cat("Unique tracts:", n_distinct(evict$GEOID), "\n")
cat("Date range:", min(evict$month_date), "to", max(evict$month_date), "\n")
```

```{r pull-renter-households}
renters_acs <- tryCatch(
  {
    get_acs(
      geography = "tract",
      variables = "B25003_003",
      state = "PA",
      county = "Philadelphia",
      year = 2023,
      geometry = FALSE
    ) %>%
      transmute(GEOID, renters = estimate)
  },
  error = function(e) {
    stop("ACS renter-household data could not be downloaded. Check your Census API key and internet connection.")
  }
)

evict <- evict %>%
  left_join(renters_acs, by = "GEOID")

summary(evict$renters)
```

```{r create-panel-features}
evict_panel <- evict %>%
  arrange(GEOID, month_date) %>%
  group_by(GEOID) %>%
  mutate(
    filings_rate_1k = (filings_2020 / renters) * 1000,
    filings_next = lead(filings_2020, 1),
    filings_next_rate_1k = lead(filings_rate_1k, 1),
    lag_1m = lag(filings_2020, 1),
    lag_2m = lag(filings_2020, 2),
    lag_3m = lag(filings_2020, 3),
    change_1m = filings_2020 - lag_1m,
    change_3m = filings_2020 - lag_3m,
    rolling_mean_3m = (lag_1m + lag_2m + lag_3m) / 3
  ) %>%
  ungroup() %>%
  mutate(
    month_index = dense_rank(month_date),
    racial_majority = factor(racial_majority),
    month_factor = factor(month_num, levels = 1:12, labels = month.abb)
  )

model_data <- evict_panel %>%
  filter(
    !is.na(filings_next),
    !is.na(lag_1m),
    !is.na(lag_2m),
    !is.na(lag_3m),
    !is.na(renters),
    renters > 0
  )

model_data %>%
  summarise(
    rows = n(),
    tracts = n_distinct(GEOID),
    avg_filing = mean(filings_next, na.rm = TRUE),
    pct_zero = mean(filings_next == 0, na.rm = TRUE)
  ) %>%
  kable(caption = "Modeling Dataset Summary", digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The modeling dataset keeps monthly observations with enough history to calculate dynamic predictors. Each row uses information known at month `t` to predict filings at month `t+1`.

# Time Split

To avoid leakage, I split the panel by time instead of using a random train-test split.

```{r create-time-split}
split_table <- tibble(
  month_date = sort(unique(model_data$month_date))
) %>%
  mutate(
    data_split = case_when(
      month_date <= ymd("2025-02-01") ~ "Train",
      month_date <= ymd("2025-08-01") ~ "Validation",
      TRUE ~ "Test"
    )
  )

model_data <- model_data %>%
  left_join(split_table, by = "month_date")

split_table %>%
  count(data_split, name = "months") %>%
  kable(caption = "Time-Based Split", digits = 0) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

train_data <- model_data %>% filter(data_split == "Train")
validation_data <- model_data %>% filter(data_split == "Validation")
test_data <- model_data %>% filter(data_split == "Test")
```

# Exploratory Check Before Modeling

```{r visualize-target}
p1 <- ggplot(train_data, aes(x = filings_next)) +
  geom_histogram(bins = 40, fill = "#3182bd", color = "white") +
  labs(
    title = "Distribution of Next-Month Filings",
    subtitle = "Training sample",
    x = "Next-month filings",
    y = "Count"
  ) +
  plotTheme

p2 <- train_data %>%
  group_by(month_date) %>%
  summarise(avg_next = mean(filings_next, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = month_date, y = avg_next)) +
  geom_line(color = "#08519c", linewidth = 1) +
  labs(
    title = "Average Next-Month Filings Over Time",
    x = "Month",
    y = "Average filings"
  ) +
  plotTheme

p1 + p2
```

The target remains right-skewed with many zeros, so count models are a natural starting point. I compare a naive baseline, a Poisson model, and a negative binomial model.

# Baseline and Count Models

```{r fit-models}
model_formula <- filings_next ~
  lag_1m +
  lag_3m +
  change_1m +
  rolling_mean_3m +
  filings_avg +
  filings_avg_prepandemic_baseline +
  racial_majority +
  month_factor +
  offset(log(renters))

poisson_model <- glm(
  formula = model_formula,
  data = train_data,
  family = poisson(link = "log")
)

nb_model <- MASS::glm.nb(
  formula = model_formula,
  data = train_data
)

cat("Poisson AIC:", AIC(poisson_model), "\n")
cat("Negative binomial AIC:", AIC(nb_model), "\n")
```

```{r score-models}
score_predictions <- function(data, split_name) {
  baseline_scored <- data %>%
    mutate(pred_next = pmax(lag_1m, 0)) %>%
    score_count_model("filings_next", "pred_next") %>%
    mutate(model = "Baseline: lag_1m", split = split_name)

  poisson_scored <- data %>%
    mutate(pred_next = predict(poisson_model, newdata = data, type = "response")) %>%
    score_count_model("filings_next", "pred_next") %>%
    mutate(model = "Poisson", split = split_name)

  nb_scored <- data %>%
    mutate(pred_next = predict(nb_model, newdata = data, type = "response")) %>%
    score_count_model("filings_next", "pred_next") %>%
    mutate(model = "Negative Binomial", split = split_name)

  bind_rows(baseline_scored, poisson_scored, nb_scored) %>%
    dplyr::select(split, model, everything())
}

validation_scores <- score_predictions(validation_data, "Validation")
test_scores <- score_predictions(test_data, "Test")

model_comparison <- bind_rows(validation_scores, test_scores)

model_comparison %>%
  mutate(across(c(MAE, RMSE, Correlation), round, 3)) %>%
  kable(caption = "Model Comparison Across Time Splits") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r visualize-model-comparison}
ggplot(model_comparison, aes(x = model, y = RMSE, fill = split)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Validation" = "#9ecae1", "Test" = "#08519c")) +
  labs(
    title = "Out-of-Sample Model Performance",
    subtitle = "Lower RMSE indicates better forecasting accuracy",
    x = NULL,
    y = "RMSE"
  ) +
  plotTheme
```

# Main Model Interpretation

I use the negative binomial model as the main specification because the EDA suggested overdispersion in filing counts.

```{r interpret-main-model}
nb_tidy <- broom::tidy(nb_model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(estimate))

nb_tidy %>%
  dplyr::select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(c(estimate, conf.low, conf.high, p.value), round, 3)) %>%
  slice(1:15) %>%
  kable(
    caption = "Negative Binomial Model: Incidence Rate Ratios",
    col.names = c("Term", "IRR", "Lower CI", "Upper CI", "p-value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The incidence rate ratios help show whether a one-unit increase in each predictor is associated with higher or lower expected filing counts in the next month, holding the other variables constant.

# Identify High-Risk Tracts

To translate predicted counts into an early-warning tool, I define high risk using the top decile of next-month filing rates per 1,000 renter households in the training sample.

```{r define-risk-threshold}
risk_threshold <- quantile(train_data$filings_next_rate_1k, probs = 0.90, na.rm = TRUE)

cat("High-risk threshold (top 10% of next-month filing rates per 1,000 renters):",
    round(risk_threshold, 3), "\n")
```

```{r score-risk}
test_scored <- test_data %>%
  mutate(
    pred_next = predict(nb_model, newdata = test_data, type = "response"),
    pred_next_rate_1k = (pred_next / renters) * 1000,
    high_risk_true = if_else(filings_next_rate_1k >= risk_threshold, 1, 0),
    high_risk_pred = if_else(pred_next_rate_1k >= risk_threshold, 1, 0)
  )

risk_scores <- score_classification(test_scored, "high_risk_true", "high_risk_pred")

risk_scores %>%
  mutate(across(everything(), round, 3)) %>%
  kable(caption = "High-Risk Classification Performance on Test Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r top-risk-tracts}
latest_forecast <- test_scored %>%
  filter(month_date == max(month_date, na.rm = TRUE)) %>%
  arrange(desc(pred_next_rate_1k)) %>%
  transmute(
    GEOID,
    racial_majority,
    current_month = month_date,
    predicted_next_filings = round(pred_next, 2),
    predicted_next_rate_1k = round(pred_next_rate_1k, 2),
    predicted_high_risk = high_risk_pred
  )

latest_forecast %>%
  slice(1:15) %>%
  kable(caption = "Top 15 Tracts by Predicted Next-Month Filing Risk") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Predicted Risk Map

```{r load-tract-geometry}
tract_shp <- list.files(
  here("data", "Census_Tracts_2010-shp"),
  pattern = "\\.shp$",
  full.names = TRUE
)[1]

philly_tracts <- st_read(tract_shp, quiet = TRUE)

geoid_field <- names(philly_tracts)[stringr::str_detect(names(philly_tracts), "GEOID")][1]

philly_tracts <- philly_tracts %>%
  mutate(GEOID = as.character(.data[[geoid_field]]))

forecast_map <- philly_tracts %>%
  left_join(latest_forecast, by = "GEOID")
```

```{r map-risk}
p_rate <- ggplot(forecast_map) +
  geom_sf(aes(fill = predicted_next_rate_1k), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "gray90",
    name = "Predicted filings\nper 1,000 renters"
  ) +
  labs(
    title = "Forecasted Eviction Filing Risk by Census Tract",
    subtitle = paste("Prediction target:", format(max(latest_forecast$current_month) %m+% months(1), "%B %Y"))
  ) +
  mapTheme

p_binary <- ggplot(forecast_map) +
  geom_sf(aes(fill = factor(predicted_high_risk)), color = NA) +
  scale_fill_manual(
    values = c("0" = "gray85", "1" = "#d7301f"),
    labels = c("0" = "Not high risk", "1" = "High risk"),
    na.translate = FALSE,
    name = "Risk class"
  ) +
  labs(
    title = "Predicted High-Risk Tracts",
    subtitle = "Tracts above the top-decile training threshold"
  ) +
  mapTheme

p_rate + p_binary
```

# Observed vs Predicted

```{r observed-vs-predicted}
ggplot(test_scored, aes(x = filings_next, y = pred_next)) +
  geom_point(alpha = 0.25, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "#d7301f", linetype = "dashed") +
  labs(
    title = "Observed vs Predicted Next-Month Filings",
    subtitle = "Negative binomial model on the test period",
    x = "Observed filings",
    y = "Predicted filings"
  ) +
  plotTheme
```

# Conclusion

This workflow turns the EDA into a tract-month forecasting pipeline. The model uses recent filing dynamics, historical tract conditions, seasonality, and renter exposure to predict next-month filings, then converts those predictions into a high-risk screening rule. In a policy setting, the final output can be used to prioritize outreach, rental assistance, or legal support in tracts where predicted filing rates are expected to spike.
