---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Chen Yitao"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the **California Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `labs/lab_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```
If there is a special character like a colon, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus) 
library(tidyverse)
library(knitr)
# Set your Census API key
census_api_key("9ca60c462873e74921e40ce793d8dfbfcd324e08", overwrite=TRUE)
census_api_key("9ca60c462873e74921e40ce793d8dfbfcd324e08", install = TRUE, overwrite = TRUE)
# Choose your state for analysis - assign it to a variable called my_state
my_state <- "CA"
```

**State Selection:** I have chosen **California** for this analysis because: I love California and I am a big fan of Los Angeles Clippers.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
ca_data <- get_acs(
  geography = "county",
  state = my_state,
  variables = c(median_hh_income = "B19013_001", total_pop = "B01003_001"),
  year = 2022,
  survey = "acs5",
  output = "wide"
)
# Clean the county names to remove state name and "County" 
ca_data
# Hint: use mutate() with str_remove()
ca_clean <- ca_data %>%
  mutate(
      county_name = str_remove(NAME, " County, California")
  )
# Display the first few rows
head(ca_clean)
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
ca_reliability <- ca_clean %>%
  mutate(
    income_moe_pct = round((median_hh_incomeM / median_hh_incomeE) * 100, 2),
    reliability = case_when(
      income_moe_pct < 5 ~ "High Confidence",
      income_moe_pct < 10 ~ "Moderate",
      TRUE ~ "Low Confidence"
    ))
print(ca_reliability, n=60)

unreliable_flag <- ca_reliability %>%
  filter(income_moe_pct > 10)
print(unreliable_flag)
# Create a summary showing count of counties in each reliability category
# Hint: use count() and mutate() to add percentages
reliability_summary <- ca_reliability %>%
  group_by(reliability) %>%
  summarize(
    counties = n(),
    avg_income = round(mean(median_hh_incomeE, na.rm = TRUE), 0)
  ) %>%
mutate(
    pct = round(counties / sum(counties) * 100, 1)
  )
reliability_summary
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage
top5_uncertainty <- unreliable_flag %>%
  arrange(desc(unreliable_flag)) %>%
  slice_head(n = 5) %>%
  select(
    county_name,
    median_hh_incomeE,
    median_hh_incomeM,
    income_moe_pct,
    reliability
  )
print(top5_uncertainty)
# Format as table with kable() - include appropriate column names and caption
top5_uncertainty %>%
  kable(
    col.names = c(
      "County",
      "Median household income (Estimate)",
      "Median household income (MOE)",
      "MOE (%)",
      "Reliability"
    ),
    caption = "Top 5 counties with highest income MOE percentage"
  )
```

**Data Quality Commentary:**

These results highlight the error and uncertainty of relying on a single data source. For the algorithm as a whole, if it depends only on median household income, it may misclassify counties with small sample sizes. In counties with large income inequality, the median may also fail to reflect the problem.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
selected_high <- ca_reliability %>%
  filter(reliability == "High Confidence") %>%
  arrange(desc(total_popE)) %>%
  slice(1)

selected_mod <- ca_reliability %>%
  filter(reliability == "Moderate") %>%
  arrange(desc(total_popE)) %>%
  slice(1)

selected_low <- ca_reliability %>%
  filter(reliability == "Low Confidence") %>%
  arrange(desc(total_popE)) %>%
  slice(1)
# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
selected_counties <- bind_rows(selected_high, selected_mod, selected_low) %>%
   select(county_name, median_hh_incomeE, income_moe_pct, reliability)
selected_counties
```

**Comment on the output:** This tool is very useful, but for areas with low data reliability, what should we do to avoid bias? I’m not sure whether I should drop those data points, or incorporate additional data sources and evaluate the issue more holistically.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter
selected_counties <- selected_counties %>%
  left_join(ca_clean %>% select(GEOID, county_name), by = "county_name")
head(selected_counties)

county_codes <- selected_counties %>%
  mutate(
    county_geoid = as.character(GEOID),              
    county_code  = str_extract(county_geoid, "\\d{3}$")  
  )
head(county_codes)

race_variables <- c(
  total_pop = "B03002_001",
  white     = "B03002_003",
  black     = "B03002_004",
  hispanic  = "B03002_012"
)

tract_053 <- get_acs(
  geography = "tract",
  variables = race_variables,
  state = my_state,
  county = "037",
  year = 2022,
  survey = "acs5",
  output = "wide"
)

tract_109 <- get_acs(
  geography = "tract",
  variables = race_variables,
  state = my_state,
  county = "103",
  year = 2022,
  survey = "acs5",
  output = "wide"
)

tract_091 <- get_acs(
  geography = "tract",
  variables = race_variables,
  state = my_state,
  county = "063",
  year = 2022,
  survey = "acs5",
  output = "wide"
)

tract_data <- bind_rows(tract_053, tract_109, tract_091)
tract_data
# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations
tract_data_pct <- tract_data %>%
  mutate(
    pct_white    = whiteE    / total_popE * 100,
    pct_black    = blackE    / total_popE * 100,
    pct_hispanic = hispanicE / total_popE * 100
  )
head(tract_data_pct)
# Add readable tract and county name columns using str_extract() or similar
tract_data_pct_readable <- tract_data_pct %>%
  mutate(
    tract_name = str_extract(NAME, "Census Tract[^;]+") %>%
      str_trim(),
    county_name = str_extract(NAME, "[^;]+ County") %>%
      str_remove(" County") %>%
      str_trim()
  )
tract_data_pct_readable
```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
top_hispanic_tract <- tract_data_pct_readable %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(GEOID, tract_name, county_name, total_popE, pct_hispanic)

top_hispanic_tract %>%
  kable(
    col.names = c("Tract GEOID", "Tract", "County", "Total Population", "% Hispanic"),
    caption = "Tract with Highest % Hispanic"
  )
# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
county_avg_demo <- tract_data_pct_readable %>%
  group_by(county_name) %>%
  summarize(
    num_tracts = n(),
    avg_pct_white = mean(pct_white, na.rm = TRUE),
    avg_pct_black = mean(pct_black, na.rm = TRUE),
    avg_pct_hispanic = mean(pct_hispanic, na.rm = TRUE)
  )

county_avg_demo %>%
  kable(
    col.names = c("County", "Number of Tracts", "Avg % White", "Avg % Black", "Avg % Hispanic/Latino"),
    caption = "Average Tract Demographics (ACS 2018–2022 5-Year)"
  )

# Create a nicely formatted table of your results using kable()
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
tract_data_moe <- tract_data_pct_readable %>%
  mutate(
    white_moe_pct = round((whiteM / whiteE) * 100, 2),
    black_moe_pct = round((blackM / blackE) * 100, 2),
    hispanic_moe_pct = round((hispanicM / hispanicE) * 100, 2),
# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement
high_moe_flag = ifelse(
      (white_moe_pct > 15) & (black_moe_pct > 15) & (hispanic_moe_pct > 15),
      "High MOE",
      "GOOD"
    )
)
tract_data_moe
# Create summary statistics showing how many tracts have data quality issues
data_quality_summary <- tract_data_moe %>%
  summarize(
    total_tracts = n(),
    high_moe_tracts = sum(high_moe_flag == "High MOE", na.rm = TRUE),
    pct_high_moe = round(high_moe_tracts / total_tracts * 100, 1)
  )
data_quality_summary
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages
pattern_summary <- tract_data_moe %>%
  group_by(high_moe_flag) %>%
  summarize(
    n_tracts = n(),
    avg_pop = round(mean(total_popE, na.rm = TRUE), 0),
    median_pop = round(median(total_popE, na.rm = TRUE), 0),
    avg_pct_white = round(mean(pct_white, na.rm = TRUE), 1),
    avg_pct_black = round(mean(pct_black, na.rm = TRUE), 1),
    avg_pct_hispanic = round(mean(pct_hispanic, na.rm = TRUE), 1),
  )

pattern_summary
# Use group_by() and summarize() to create this comparison
# Create a professional table showing the patterns
pattern_summary %>%
  kable(
    col.names = c(
      "Data Quality Group",
      "Number of Tracts",
      "Avg Population",
      "Median Population",
      "Avg % White",
      "Avg % Black",
      "Avg % Hispanic/Latino"
    ),
    caption = "Patterns in Data Quality: Comparing Tracts with High MOE vs OK"
  )
```

**Pattern Analysis:** Using the rule “if the MOE of any racial/ethnic group is greater than 15%, classify the tract as High MOE,” all census tracts are labeled as High MOE. Based on what I observed, this happens largely because the Black population count is often small, while the margin of error is relatively large, which makes the Black MOE percentage high for many tracts.

I then changed the rule to “a tract is High MOE only if the MOE of all groups is greater than 15%.” Under this stricter definition, 86% of tracts are still classified as High MOE, which suggests that uncertainty in the Black estimate is not the only reason for the high High-MOE rate.

In the pattern analysis, High MOE tracts have lower average and median total population, a lower share of Hispanic/Latino residents, and a higher share of Black residents. Keeping the same approach, I raised the High MOE threshold to 20%. At this threshold, the shares of High MOE and Good tracts are close to a 50–50 split, but the pattern analysis still shows a similar structure. Therefore, I suspect that tracts with smaller populations and a higher Black share tend to have higher MOE and worse data quality.


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?
2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?
3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?
4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

**Executive Summary:**

I found the clearest and most reliable systematic pattern to be that, for sample-based ACS data, statistical reliability tends to decrease as the unit of analysis becomes smaller. For example, when analyzing median household income at the county level, 70.7% of California counties fall into the High Confidence category. However, when the analysis is refined to the census tract level, MOE percentages generally increase. In addition, for Los Angeles County—the most populous county—race/ethnicity estimates are relatively reliable at the county level, with low MOE. But once the analysis is further broken down into individual tracts within the county, many tracts are still classified as High MOE, likely because each tract has a smaller sample size and smaller subgroup counts.

Based on my results, the communities most vulnerable to algorithmic bias are small-population tracts, because their ACS estimates have higher uncertainty and are more likely to be flagged as High MOE. In my tract-level pattern analysis, High MOE tracts also tend to have a higher Black share and a lower Hispanic/Latino share, while the White share differs very little between groups. However, I cannot claim these demographic patterns are direct causal drivers of bias risk. A more defensible interpretation is that algorithmic bias risk increases where the underlying inputs are statistically less reliable—often due to smaller effective sample sizes and smaller subgroup counts at the tract level—so algorithmic bias may be more prone to be seen in those areas.

As the geographic scale becomes smaller (from counties to census tracts), uncertainty in ACS estimates generally increases. This is mainly because sample-based surveys have smaller effective sample sizes in small geographic units, and because some racial/ethnic groups have smaller absolute counts within a tract, which can substantially increase MOE percentages. At the same time, my results show that High MOE is more common in tracts with smaller populations and demographic compositions that differ from the overall average, suggesting that data quality issues and potential bias risk may be more concentrated in specific communities. I also found that ACS data quality may vary across states and counties; for example, Pennsylvania has no counties with High MOE (greater than 10%). Therefore, cross-regional conclusions should be confirmed by running the same tests in additional locations.

When the Department encounters areas with high MOE, it should flag them as unreliable areas rather than directly using ACS data. In practice, the Department can set a clear “High MOE” threshold. When MOE is too high, the data should not be used for fully automated decisions. Instead, the Department should use more robust approaches, such as merging adjacent tracts, analyzing at a larger geographic scale, or validating results with multiple data sources. If high-MOE areas might be underestimated or overlooked by the algorithm, the Department should also add safeguards, such as human review, a minimum service floor, or targeted policy support.

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category
county_reliability_table <- ca_reliability %>%
  select(
    county_name,
    median_hh_incomeE,
    median_hh_incomeM,
    income_moe_pct,
    reliability
  )
# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"
county_reliability_table %>%
 mutate(
    recommendation = case_when(
      reliability == "High Confidence" ~ "Safe for algorithmic decisions",
      reliability == "Moderate" ~ "Use with caution - monitor outcomes",
      reliability == "Low Confidence" ~ "Requires manual review or additional data",
      TRUE ~ "Unknown"
    )) %>%
arrange(desc(income_moe_pct)) %>%
# Format as a professional table with kable()
kable(
    col.names = c(
      "County",
      "Median HH Income (Estimate)",
      "Median HH Income (MOE)",
      "Income MOE (%)",
      "Reliability",
      "Recommendation"
    ),
    caption = "Decision Framework: Algorithm Recommendations by County Data Reliability (ACS 2022 5-Year, California)",
    format.args = list(big.mark = ",")  # Add commas to large numbers
  )
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** 
```{r high confidence}
high_conf_counties <- ca_reliability %>%
  filter(reliability == "High Confidence") %>%
  arrange(income_moe_pct) %>%         
  select(county_name, median_hh_incomeE, income_moe_pct)
high_conf_counties %>% pull(county_name)
```
These counties have relatively low MOE percentages, which indicates higher statistical reliability in the ACS estimates. Therefore, these counties are considered “Safe for algorithmic decisions.”

2. **Counties requiring additional oversight:** 
```{r moderate}
moderate_counties <- ca_reliability %>%
  filter(reliability == "Moderate") %>%
  arrange(income_moe_pct) %>%         
  select(county_name, median_hh_incomeE, income_moe_pct)
moderate_counties %>% pull(county_name)
```
Moderate Confidence counties (MOE% 5–10%) are still usable, but their uncertainty requires additional oversight. The Department should validate key inputs using multiple data sources. For counties near decision thresholds, human review should be applied. In addition, for these uncertain counties, decisions should not rely on data alone; instead, they should incorporate local conditions and combine real-world context with the data when making decisions.

3. **Counties needing alternative approaches:** 
```{r low confidence}
low_conf_counties <- ca_reliability %>%
  filter(reliability == "Low Confidence") %>%
  arrange(income_moe_pct) %>%         
  select(county_name, median_hh_incomeE, income_moe_pct)
low_conf_counties %>% pull(county_name)
```
For Low Confidence counties, the Department should avoid direct algorithmic use of the ACS estimate. First, require human review for cases where this variable could change the decision, and validate the ACS values with other sources. Second, decisions should be based on multiple data sources rather than ACS alone, and the Department should document which source is used when ACS reliability is low. Finally, monitor algorithm outcomes monthly or quarterly for unusual shifts, and recompute reliability categories annually when new ACS data are released.

## Questions for Further Investigation

1. Are High MOE tracts spatially clustered within each selected county, and do they align with specific place types such as low-density areas or mountainous regions?
2. How stable are these MOE patterns over time?
3. Beyond race/ethnicity, which demographic factors are most associated with high uncertainty. For example poverty rate, age structure, housing tenure,or educational?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on [date]

**Reproducibility:** 
- All analysis conducted in R version 4.5.2
- Census API key required for replication 9ca60c462873e74921e40ce793d8dfbfcd324e08
- Complete code and documentation available at: https://cyt1014.github.io/cpln-5920-sp26-student-portfolio-template/

**Methodology Notes:**
I selected California (CA) as the study state. For county selection, I chose the most populous county within each reliability category (High, Moderate, Low) to avoid cases with very few census tracts. For tract-level classification, I defined a tract as “Low Confidence” only when the MOE% for all three race/ethnicity variables exceeded 15% (White, Black, and Hispanic all had MOE% > 15%).

**Limitations:**
This analysis is limited to California. The investigation of race/ethnicity MOE patterns is based on only three large-population counties, so the findings may not be generalizable. In addition, the analysis uses ACS data from a single year/period, which may not capture recent changes.

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/labs/lab_1/your_file_name.html`